Cultural Transmission and the Social Contexts of Technological Transitions during the Late Paleolithic of Korea

### 2023-04-22 analysis/paper/paper.qmd
editing the conclusion

In addition to correlation coefficients of point attributes, we measured coefficients of variation (CVs) to quantify variability in individual attributes. @fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv`% to `r max_cv`%. Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for Korean artifacts is higher. The CVs for body length (BL) and maximum length (ML) have the lowest values. The low CV and narrow confidence intervals for maximum length indicate that this dimension is highly standardized relative to the others. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes. Overall, with all but one attribute having CV values above our threshold value, our results suggest that the transmission of stemmed points was mostly influenced by guided variation.

@@

The most striking differences in both correlation coefficients and CV values are in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material). Raw materials likely had a substantial impact on stemmed points' shape variance. We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not exactly match our expectations because of how influential raw materials are on CV values. For example, artifacts made from acidic quartzite have the highest correlations, but their CV values are also much higher with wider confidence intervals than all other raw materials. These complex results might be related to the relatively unpredictable knapping quality of quartzite. Large grains within quartzite can redirect knapping forces in unexpected directions, resulting in unintended flake removals. This makes it harder for the knapper to repeatedly produce the same size and shape end product. Variation in raw material nodule size may also contribute to this pattern, for example quartzite nodules may have been found in a much wider range of sizes than nodules of other raw materials. We could not strongly argue this here due to the small sample size of quartzite to generalize this pattern, but this hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability. Further analysis of raw material properties and availability is needed to make robust conclusions about their influence here.

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka. It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also found a slight decrease in CV values for the four attributes. Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission. Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. Another possibility is that the cooler climate might result in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources, which would have concentrated them in fewer predictable locations. In either case, these conditions might lead to larger networks of individuals interacting with one another and an increased ability to learn from experienced social teachers or successful models.

In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive, exclusive cultural transmission biases in tool-making in different assemblages. Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3. Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period. One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did]. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.

We concede that our work is rather speculative and provides only a limited explanation of the social contexts surrounding the spread of stemmed points. Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span. In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current], but our research sites lacked sufficient information to utilize this method. In our future study, we will further investigate how to quantify multiple variables and conduct simulations to develop a logical bridge between our primary approach borrowed from Bettinger and Eerkens [-@bettinger1999point] and social contexts for the appearance of stemmed points. We acknowledge that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e. the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional]. It is our assumption that the majority of stemmed points used in our research have retained their original shape, given that retouch is rare on the tip of stemmed points. In our study, we only considered two types of biases as social contexts without considering other factors that impact on morphology of stemmed points such as tool efficiency and manufacturing economization. Mainly due to limited empirical evidence to explore a multiplicity of conjectural scenarios, we decided to focus on our primary goal, understanding the general trajectory of technological transition. In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points. By so doing, we anticipate achieving a more exhaustive understanding of the morphology of stemmed points and the mechanisms behind their transmission.

>>>>

In addition to correlation coefficients of point attributes, we measured coefficients of variation (CVs) to quantify variability in individual attributes. @fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv`% to `r max_cv`%. Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for Korean artifacts is higher. The CVs for body length (BL) and maximum length (ML) have the lowest values. The low CV and narrow confidence intervals for maximum length indicate that this dimension is highly standardized relative to the others. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes. Overall, with all but one attribute having CV values above our threshold value at this level of aggregation of assemblages, our results suggest that the transmission of stemmed points was mostly influenced by guided variation.

@@

The most striking differences in both correlation coefficients and CV values are in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material). Raw materials likely had a substantial impact on stemmed points' shape variance. We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not exactly match our expectations because of how influential raw materials are on CV values. For example, artifacts made from acidic quartzite have the highest correlations, but their CV values are also much higher with wider confidence intervals than all other raw materials. These complex results might be related to the relatively unpredictable knapping quality of quartzite. Large grains within quartzite can redirect knapping forces in unexpected directions, resulting in unintended flake removals. This makes it harder for the knapper to repeatedly produce the same size and shape end product. Variation in raw material nodule size may also contribute to this pattern, for example quartzite nodules may have been found in a much wider range of sizes than nodules of other raw materials. The small sample size of quartzite artefacts in our study limits how far we can generalize this pattern, but this hints at the possibility that the physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability. Further analysis of raw material properties and availability is needed to make robust conclusions about their influence here.

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka. It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also found a slight decrease in CV values for the four attributes. Those results can be interpreted as a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skill transmission. In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive, exclusive cultural transmission biases in tool-making in different assemblages. Correlation coefficient values across the assemblages confirm the overall chronological trend indicating indirect bias in Phase 3. Raw material choice appears to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period.

One possibility is that these more easily manageable raw materials were part of an adaptive shift during the Last Glacial Maximum, where perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, changes in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did]. Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. Another possibility is that the cooler climate might have resulted in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources <!-- citation needed, or delete this sentence --> , which would have concentrated them in fewer predictable locations. In either case, these conditions likely led to larger networks of individuals interacting with one another and an increased opportunities to learn from experienced teachers or successful models. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.

A key difference between our work and Bettinger and Eerkens and Garvey is the time span that we are attempting to infer cultural transmission for. For Bettinger and Eerkens and Garvey the time span is in the order of 100s of years or dozens of human generations, while our study spans 1000s of years and 100s of human generations.  In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current].

Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span. but our research sites lacked sufficient information to utilize this method. In our future study, we will further investigate how to quantify multiple variables and conduct simulations to develop a logical bridge between our primary approach borrowed from Bettinger and Eerkens [-@bettinger1999point] and social contexts for the appearance of stemmed points. We acknowledge that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e. the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional]. It is our assumption that the majority of stemmed points used in our research have retained their original shape, given that retouch is rare on the tip of stemmed points. In our study, we only considered two types of biases as social contexts without considering other factors that impact on morphology of stemmed points such as tool efficiency and manufacturing economization. Mainly due to limited empirical evidence to explore a multiplicity of conjectural scenarios, we decided to focus on our primary goal, understanding the general trajectory of technological transition. In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points. By so doing, we anticipate achieving a more exhaustive understanding of the morphology of stemmed points and the mechanisms behind their transmission.
